---
title: "session2"
author: "jolynecujoh233"
date: "2026-03-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
# 设置R Markdown代码块的全局参数  Set global options for R Markdown code chunks
knitr::opts_chunk$set(echo = TRUE, # 显示代码 / show code in output
                      warning = FALSE, # 不显示警告 / suppress warnings
                      message = FALSE, # 不显示package提示 / suppress messages
                      fig.width = 9,   # 图宽 / figure width
                      fig.height = 5)  # 图高 / figure height
options(max.print = 80) # 控制打印输出的最大行数
```


```{r packages}
library(tidyverse)
library(ggtext)
library(patchwork)
library(ggrepel)
library(scales)

# 设置默认绘图主题
theme_set(theme_minimal())

# 定义颜色调色板
pal <- c(
  nus_blue   = "#003D7C",
  blue       = "#2E75B6",
  nus_orange = "#EF7C00",
  ice        = "#D6E8F7",
  grey       = "#6B7D8C"
)
```

```{r install, eval=FALSE}
install.packages(c("ggtext", "patchwork", "ggrepel", "scales"))
```

```{r load-files}
# List all files in the data directory
files <- list.files("D:/2/Data/res", full.names = TRUE)
```

```{r read}
# Read all files and stack them into one data frame
residents_raw <- read_csv(files)
```

```{r rename}
# 重命名变量
residents <- residents_raw %>%
  rename(
    planning_area = PA,
    subzone       = SZ,
    age_group     = AG,
    sex           = Sex,
    dwelling_type = TOD,
    residents     = Pop,
    year          = Time
  )
```

```{r recode-90}
# 修正数据值 修改年龄组标签
residents <- residents %>% mutate(age_group = 
                              recode(age_group,
                                '90_and_Over' = '90_and_over'
                                    )
                                  )
```


```{r year-focus}
y_latest <- max(residents$year)
y_latest
```

```{r year-comparison}
# 计算2000年与最新年份的人口变化
pa_change <- residents %>%
  filter(year %in% c(2000, y_latest)) %>% # 只保留2000和最新年份
  group_by(planning_area, year) %>% # 按规划区+年份分组
  summarise(total = sum(residents, na.rm = TRUE), .groups = "drop") %>% # 计算每个区域的总人口
  pivot_wider(           # 宽表格式：2000和2025变成两列
    names_from   = year,
    values_from  = total,
    names_prefix = "y"
  ) %>%
  mutate(     # 计算变化
    change     = y2025 - y2000,
    pct_change = 100 * change / y2000,
    direction  = if_else(change >= 0, "Growth", "Decline")
  ) %>%
  filter(!is.na(y2000), y2000 > 0, y2025 > 0) %>%  # 删除无效数据
  arrange(desc(change))

pa_change %>% head(10)
```


```{r simplify-dwelling}
# 简化住房类型
residents <- residents %>%
  mutate(
    dwelling_broad = case_when(
      str_detect(dwelling_type, "HDB 1")            ~ "HDB Small (1–2 Room)",
      str_detect(dwelling_type, "HDB 3")            ~ "HDB 3-Room",
      str_detect(dwelling_type, "HDB 4")            ~ "HDB 4-Room",
      str_detect(dwelling_type, "HDB 5|Executive")  ~ "HDB Large (5-Room+)",
      str_detect(dwelling_type, "Condominium")      ~ "Condo / Apartment",
      str_detect(dwelling_type, "Landed")           ~ "Landed",
      str_detect(dwelling_type, "Other")            ~ "Other / HUDC",
      str_detect(dwelling_type, "HUDC")             ~ "Other / HUDC",
      TRUE                                          ~ "NA"
      # The last one (else) should not happen because above we covered all options
    )
  )

# Like always, we should check the output just in case
residents %>%
  count(dwelling_broad, dwelling_type) %>%
  arrange(dwelling_broad)
```

## Section 6: Visualising the answer

### Plot 1: Top growing and declining areas (bar chart)

```{r growth-bar-step1}
growth_subset <- bind_rows(
  pa_change %>% slice_max(change, n = 10),
  pa_change %>% slice_min(change, n = 10)
) %>%
  mutate(planning_area = fct_reorder(planning_area, change))

# Step 1: plain bar
ggplot(growth_subset, aes(x = change, y = planning_area)) +
  geom_col()
```

```{r growth-bar-final}
ggplot(growth_subset, aes(x = change, y = planning_area, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Growth" = pal[["nus_blue"]], "Decline" = pal[["nus_orange"]])) +
  labs(
    title    = "Resident population change by planning area, 2000–2025",
    subtitle = "Top 10 growing and top 10 declining areas",
    x        = "Change in residents (2000 -> 2025)",
    y        = NULL,
    fill     = NULL,
    caption  = "Source: Singapore Department of Statistics"
  ) +
  theme(legend.position = "top")
```


### Plot 2: Dumbbell chart: 2000 vs latest population

```{r dumbbell-prep}
dumbbell_data <- pa_change %>%
  filter(y2025 >= 20000) %>%
  mutate(planning_area = fct_reorder(planning_area, y2025))

dumbbell_data
```

#### Step 1: segments only


```{r dumbbell-step1}
ggplot(dumbbell_data, aes(y = planning_area)) +
  geom_segment(
    aes(x = y2000, xend = y2025, yend = planning_area),
    colour = "grey70",
    linewidth = 0.9
  )
```

#### Step 2: add coloured endpoints

```{r dumbbell-step2}
ggplot(dumbbell_data, aes(y = planning_area)) +
  geom_segment(
    aes(x = y2000, xend = y2025, yend = planning_area),
    colour = "grey70", linewidth = 0.9
  ) +
  geom_point(aes(x = y2000), colour = pal["nus_blue"],   size = 3.5) +
  geom_point(aes(x = y2025), colour = pal["nus_orange"], size = 3.5)
```

#### Step 3: add axis formatting and a colour-coded subtitle.

```{r dumbbell-final}
ggplot(dumbbell_data, aes(y = planning_area)) +
  geom_segment(
    aes(x = y2000, xend = y2025, yend = planning_area),
    colour = "grey70", linewidth = 0.9
  ) +
  geom_point(aes(x = y2000), colour = pal["nus_blue"],   size = 3.5) +
  geom_point(aes(x = y2025), colour = pal["nus_orange"], size = 3.5) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title    = "Resident population: 2000 vs now",
    subtitle = "<span style='color:#003D7C'>&#9679; 2000</span>  &nbsp;&nbsp;  <span style='color:#EF7C00'>&#9679; 2025</span>",
    x        = "Residents",
    y        = NULL,
    caption  = "Source: Singapore Department of Statistics"
  ) +
  theme(plot.subtitle = ggtext::element_markdown())
```

### Plot 3: Dwelling type composition


```{r dwelling-shares}
dwelling_shares <- residents %>%
  filter(year == y_latest,
         planning_area %in% dumbbell_data$planning_area) %>%
  group_by(planning_area, dwelling_broad) %>%
  summarise(total = sum(residents, na.rm = TRUE), .groups = "drop") %>%
  group_by(planning_area) %>%
  mutate(share = total / sum(total)) %>%
  ungroup() %>%
  mutate(planning_area = fct_reorder(
    planning_area,
    share * (dwelling_broad == "HDB Large (5-Room+)"),
    .fun = sum
  ))

# dwelling_shares %>% head(10)
glimpse(dwelling_shares)
```

#### Step 1: basic stacked bar


```{r dwelling-step1}
ggplot(dwelling_shares, aes(x = share, y = planning_area, fill = dwelling_broad)) +
  geom_col(position = "stack")
```
#### Step 2: percentage x-axis, particular palette, and legend at bottom

```{r mixed-palette}
palette_dwelling <- c(
  "HDB Small (1–2 Room)" = "#D6E8F7",
  "HDB 3-Room"           = "#82B9E0",
  "HDB 4-Room"           = "#2E75B6",
  "HDB Large (5-Room+)"  = "#1A3A5C",
  "Condo / Apartment"    = "#E8703A",
  "Landed"               = "#C04010",
  "Other / HUDC"         = "#9E9E9E"
)
```


```{r dwelling-final}
ggplot(dwelling_shares, aes(x = share, y = planning_area, fill = dwelling_broad)) +
  geom_col(position = "stack") +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(values = palette_dwelling) +
  labs(
    title   = "Dwelling type composition by planning area",
    x       = "Share of residents",
    y       = NULL,
    fill    = "Dwelling type",
    caption = "Source: Singapore Department of Statistics"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

### Plot 4: Age structure

```{r age-structure}
age_order <- c("0_to_4","5_to_9","10_to_14","15_to_19","20_to_24",
               "25_to_29","30_to_34","35_to_39","40_to_44","45_to_49",
               "50_to_54","55_to_59","60_to_64","65_to_69","70_to_74",
               "75_to_79","80_to_84","85_to_89","90_and_over")

age_profiles <- residents %>%
  filter(
    year == y_latest,
    planning_area %in% c("Punggol", "Sengkang", "Toa Payoh", "Queenstown")
  ) %>%
  group_by(planning_area, age_group) %>%
  summarise(total = sum(residents, na.rm = TRUE), .groups = "drop") %>%
  group_by(planning_area) %>%
  mutate(
    share     = total / sum(total),
    age_group = factor(age_group, levels = age_order)
  ) %>%
  ungroup()

age_profiles
```

#### Step 1: basic faceted bar

```{r age-step1}
ggplot(age_profiles, aes(x = age_group, y = share)) +
  geom_col() +
  facet_wrap(~planning_area, ncol = 2) +
  coord_flip()
```

#### Step 2: percentage axis, fill by area, labels

```{r age-final}
ggplot(age_profiles, aes(x = age_group, y = share, fill = planning_area)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~planning_area, ncol = 2) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set1") +
  coord_flip() +
  labs(
    title    = "Age structure of residents",
    subtitle = "Contrasting new towns vs mature estates",
    x        = "Age group",
    y        = "Share of residents",
    caption  = "Source: Singapore Department of Statistics"
  )
```


### Plot 5: Dumbbell of condominium share (2000 vs now)

```{r condo-dumbbell-prep}
condo_change <- residents %>%
  filter(year %in% c(2000, y_latest)) %>%
  group_by(planning_area, year) %>%
  mutate(pa_total = sum(residents, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(str_detect(dwelling_type, "Condominium")) %>%
  group_by(planning_area, year, pa_total) %>%
  summarise(condo = sum(residents, na.rm = TRUE), .groups = "drop") %>%
  mutate(condo_share = condo / pa_total) %>%
  select(planning_area, year, condo_share) %>%
  pivot_wider(
    names_from   = year,
    values_from  = condo_share,
    names_prefix = "y"
  ) %>%
  filter(!is.na(y2000), !is.na(y2025)) %>%
  mutate(
    change        = y2025 - y2000,
    planning_area = fct_reorder(planning_area, y2025)
  ) %>%
  filter(pmax(y2000, y2025) > 0.05) %>%
  arrange(desc(y2025))
```


```{r condo-dumbbell-final}
ggplot(condo_change, aes(y = planning_area)) +
  geom_segment(
    aes(x = y2000, xend = y2025, yend = planning_area),
    colour = "grey70", linewidth = 0.9
  ) +
  geom_point(aes(x = y2000), colour = pal["nus_blue"],   size = 3.5) +
  geom_point(aes(x = y2025), colour = pal["nus_orange"], size = 3.5) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "Condominium share of residents: 2000 vs now",
    subtitle = "<span style='color:#003D7C'>&#9679; 2000</span>  &nbsp;&nbsp;  <span style='color:#EF7C00'>&#9679; now</span>",
    x        = "Share of residents in condominiums",
    y        = NULL,
    caption  = "Source: Singapore Department of Statistics"
  ) +
  theme(plot.subtitle = ggtext::element_markdown())
```

------------------------------------------------------------------------

## Answering the research question

```{r answer-summary}
pa_change %>%
  slice_max(change, n = 5) %>%
  select(planning_area, y2000, y2025, change, pct_change) %>%
  mutate() %>%
  knitr::kable(
    col.names   = c("Planning area", "Pop 2000", "Pop 2025", "Change", "% change"),
    format.args = list(big.mark = ",")
  )
```

```{r answer-summary2}
pa_change %>%
  slice_max(change, n = 5) %>%
  select(planning_area, y2000, y2025, change, pct_change) %>%
  mutate(across(where(is.numeric), round)) %>%
  knitr::kable(
    col.names   = c("Planning area", "Pop 2000", "Pop 2025", "Change", "% change"),
    format.args = list(big.mark = ",")
  )
```
